---
title: "Exploratory analysis script"
date: "2024-03-20"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyverse)
```


Load the data.
```{r}
#load data
hrs <- readRDS(here("data", "processed-data", "hrs_cognition_final.rds"))
```

# Data exploration 
```{r}
glimpse(hrs)
dim(hrs)
hrs %>% summarise(
  n_obs = n(),
  n_people = n_distinct(HHIDPN),
  n_waves = n_distinct(STUDYYR)
)
```



# Data exploration through tables

Showing a bit of code to produce and save a summary table.


```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.


# Data exploration through figures

Histogram plots for the continuous outcomes. 

Height first.

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures","height-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures","weight-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results", "figures", "height-weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results", "figures", "height-weight-stratified.png")
ggsave(filename = figure_file, plot=p4) 
```



# Notes



