---
title: "Exploratory analysis script"
date: "2026-03-20"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(scales)
library(tidyverse)
library(gtsummary)
```

Load the data.
```{r}
#load data
hrs_final <- readRDS(here("data", "processed-data", "hrs_final_long.rds"))
```

# Data exploration 
## quick summary of the data
```{r}
summary(hrs_final)
```

## P1. Sample size per wave
```{r}
# This figure shows the number of observations in each survey wave from 1996 to 2020.
p1 <- hrs_final %>%
  count(STUDYYR) %>%
  ggplot(aes(x = factor(STUDYYR), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = n), vjust = -0.5, size = 3) +
  labs(title = "Sample Size per Wave",
       x = "Study Year", y = "Number of Observations") +
  theme_minimal()
print(p1)

ggsave(here("results", "figures", "p1_sample_size_per_wave.png"), 
       plot = p1, width = 8, height = 5, dpi = 300)
```
## P2. Cognitive function distribution by wave
```{r}
# Proportion of Normal, CIND, and Demented individuals at each wave. An increasing proportion of dementia over time is expected.
cogfunction_table <- hrs_final %>%
  filter(!is.na(COGFUNCTION)) %>%
  count(STUDYYR, COGFUNCTION) %>%
  group_by(STUDYYR) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

p2 <- cogfunction_table %>%
  ggplot(aes(x = factor(STUDYYR), y = pct, fill = COGFUNCTION)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Normal"   = "#2196F3",
                               "CIND"     = "#FF9800",
                               "Demented" = "#F44336")) +
  labs(title = "Cognitive Function Distribution by Wave",
       x = "Study Year", y = "Percentage (%)",
       fill = "Cognitive Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("results", "figures", "p2_cogfunction_by_wave.png"),
       plot = p2, width = 10, height = 6, dpi = 300)
```

## Table 1. Baseline characteristics of the study population
This table shows the baseline characteristics of all eligible individuals at the 1996 wave (baseline).
```{r}
# Table 1: Baseline characteristics of the overall study population
table1 <- hrs_final %>%
  filter(STUDYYR == 1996) %>%
  distinct(HHIDPN, .keep_all = TRUE) %>%
  select(
    BASELINE_AGE, BASELINE_BMI, LN_WEALTH, LN_HITOT, CHRONIC_N,
    RAGENDER, RACE_ETH3, RAEDUC, HAS_INS,
    EMP_STATUS, SMOKESTATUS, R_DEPRES, BASELINE_BMI_CAT
  ) %>%
  mutate(CHRONIC_N = as.numeric(CHRONIC_N)) %>%
  tbl_summary(
    type = list(CHRONIC_N ~ "continuous"),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      BASELINE_AGE     ~ "Age at baseline (years)",
      BASELINE_BMI     ~ "BMI at baseline (kg/m²)",
      LN_WEALTH        ~ "Log household wealth",
      LN_HITOT         ~ "Log total income",
      CHRONIC_N        ~ "Number of chronic conditions",
      RAGENDER         ~ "Gender",
      RACE_ETH3        ~ "Race/Ethnicity",
      RAEDUC           ~ "Education",
      HAS_INS          ~ "Health insurance",
      EMP_STATUS       ~ "Employment status",
      SMOKESTATUS      ~ "Smoking status",
      R_DEPRES         ~ "Depression",
      BASELINE_BMI_CAT ~ "BMI category"
    ),
    missing = "no"
  ) %>%
  bold_labels() %>%
  modify_caption("**Table 1. Baseline Characteristics of the Study Population**")

table1

# Save
table1 %>%
  as_tibble() %>%
  write_csv(here("results", "tables", "t1_baseline.csv"))
```

## P3. Dementia prevalence by wave
This figure shows the prevalence of dementia at each wave. We expect to see an increasing trend in dementia prevalence over time as the cohort ages.
```{r}
dementia_incidence <- hrs_final %>%
  filter(!is.na(COGFUNCTION)) %>%
  group_by(STUDYYR) %>%
  summarise(
    n_total    = n(),
    n_demented = sum(COGFUNCTION == "Demented"),
    rate       = round(n_demented / n_total * 100, 1),
    .groups    = "drop"
  )

p_cog2 <- dementia_incidence %>%
  ggplot(aes(x = factor(STUDYYR), y = rate)) +
  geom_line(aes(group = 1), color = "#F44336", linewidth = 1) +
  geom_point(color = "#F44336", size = 3) +
  geom_text(aes(label = paste0(rate, "%")),
            vjust = -0.8, size = 3) +
  labs(title = "Dementia Prevalence Rate by Wave",
       x = "Study Year", y = "Prevalence (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("results", "figures", "p3_dementia_prevalence_by_wave.png"),
       plot = p_cog2, width = 10, height = 6, dpi = 300)
```

## Exposure Variable Distribution
## P4. Distribution of BMI Coefficient of Variation (CV%) across wave windows
This figure shows the distribution of BMI CV% across the three wave windows (W3, W6, W9). We expect to see a right-skewed distribution with most individuals having low BMI variability and a tail of individuals with higher variability.
```{r}
p_exp1 <- hrs_final %>%
  distinct(HHIDPN, .keep_all = TRUE) %>%
  select(HHIDPN, BMI_CV_W3, BMI_CV_W6, BMI_CV_W9) %>%
  pivot_longer(-HHIDPN,
               names_to  = "window",
               values_to = "BMI_CV") %>%
  mutate(window = case_when(
    window == "BMI_CV_W3" ~ "W3 (3 waves)",
    window == "BMI_CV_W6" ~ "W6 (6 waves)",
    window == "BMI_CV_W9" ~ "W9 (9 waves)"
  )) %>%
  filter(!is.na(BMI_CV)) %>%
  ggplot(aes(x = BMI_CV, fill = window, color = window)) +
  geom_density(alpha = 0.4) +
  labs(title = "Distribution of BMI Coefficient of Variation (CV%)",
       subtitle = "Across W3, W6, and W9 wave windows",
       x = "BMI CV (%)", y = "Density",
       fill = "Wave Window", color = "Wave Window") +
  theme_minimal()

ggsave(here("results", "figures", "p4_bmi_cv_distribution.png"),
       plot = p_exp1, width = 10, height = 6, dpi = 300)
```

```{r}
p_exp2 <- hrs_final %>%
  distinct(HHIDPN, .keep_all = TRUE) %>%
  select(HHIDPN, starts_with("BMI_CV_W")) %>%
  pivot_longer(-HHIDPN,
               names_to  = "window",
               values_to = "BMI_CV") %>%
  mutate(window = str_remove(window, "BMI_CV_")) %>%
  filter(!is.na(BMI_CV)) %>%
  ggplot(aes(x = window, y = BMI_CV, fill = window)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(title = "BMI CV(%) by Wave Window (W3-W9)",
       x = "Wave Window", y = "BMI CV (%)") +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(here("results", "figures", "p5_bmi_cv_boxplot.png"),
       plot = p_exp2, width = 10, height = 6, dpi = 300)

```

```{r}
# Check BMI CV outliers
hrs_final %>%
  distinct(HHIDPN, .keep_all = TRUE) %>%
  select(starts_with("BMI_CV_W")) %>%
  summary()

# Check how many people have BMI_CV > 20% (extreme outliers)
hrs_final %>%
  distinct(HHIDPN, .keep_all = TRUE) %>%
  summarise(
    n_cv_w3_gt20 = sum(BMI_CV_W3 > 20, na.rm = TRUE),
    n_cv_w6_gt20 = sum(BMI_CV_W6 > 20, na.rm = TRUE),
    n_cv_w9_gt20 = sum(BMI_CV_W9 > 20, na.rm = TRUE),
    pct_w3 = round(mean(BMI_CV_W3 > 20, na.rm = TRUE) * 100, 2),
    pct_w6 = round(mean(BMI_CV_W6 > 20, na.rm = TRUE) * 100, 2),
    pct_w9 = round(mean(BMI_CV_W9 > 20, na.rm = TRUE) * 100, 2)
  )

# Check what BMI values these extreme outliers have
hrs_final %>%
  filter(BMI_CV_W9 > 20) %>%
  distinct(HHIDPN, .keep_all = TRUE) %>%
  select(HHIDPN, BASELINE_BMI, BMI_CV_W9, BMI_MEAN_W9, BMI_SD_W9) %>%
  arrange(desc(BMI_CV_W9)) %>%
  head(20)
```


# Notes



