---
title: "Cleaning and processing code for HRS data"
author: "Liang Lai"
date: "2026-02-27"
output: html_document
---

# Setup
Load needed packages. make sure they are installed.
```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) 
library(tidyverse)
library(haven) 
library(stringr)
library(zoo)
```

```{r}
# Helper function to track sample size
check_n <- function(data) {
  data %>%
    summarise(
      n_obs = n(),
      n_people = n_distinct(HHIDPN),
      n_waves = n_distinct(STUDYYR)
    )
}
```
# Data loading
Link to the HRS data: <https://hrsdata.isr.umich.edu/data-products/long-format-data-rand-hrs-longitudinal-file-2020>. 
In this link, you can find the long format data files (rlong, hlong, rwide) that we will use for this project. You can also find the codebook and documentation for these datasets in the same link.

Link to the lange weir cognition data: <https://hrsdata.isr.umich.edu/data-products/langa-weir-classification-cognitive-function-1995-2022>. This dataset contains the cognitive classification variable (cogimp) that we will use to determine dementia status in our analysis. We will merge this cognition data with the main HRS data later in the cleaning process. However, the cognition data is in wide format and we will need to reshape it to long format before merging.
```{r}
# load HRS data (long format)  
rlong <- read_sas(here("data", "raw-data", "rlong_table.sas7bdat"))
hlong <- read_sas(here("data", "raw-data", "hlong_table.sas7bdat"))
rwide <- read_sas(here("data", "raw-data", "rwide_table.sas7bdat"))

# load lange weir data 
lange_weir_raw <- read_dta(here("data","raw-data","cogfinalimp_9522wide.dta"))
```

```{r}
# check all overlapping columns
overlap_cols <- c(
  intersect(names(rlong), names(hlong)),
  intersect(names(rlong), names(rwide)),
  intersect(names(hlong), names(rwide))
) %>% 
  unique() %>%
  setdiff(c("HHIDPN", "WAVE_NUMBER"))

# Drop overlapping columns from hlong and rwide, then merge
hrs_long <- rlong %>%
  full_join(hlong %>% select(-any_of(overlap_cols)), by = c("HHIDPN", "WAVE_NUMBER")) %>%
  left_join(rwide %>% select(-any_of(overlap_cols)), by = "HHIDPN") %>%
  arrange(HHIDPN, WAVE_NUMBER)

rm(rlong, hlong, rwide)
# Check sample size
check_n(hrs_long)
```

```{r}
#  Sample code to select 1000 random rows from the raw data and save as a new RDS file for testing and sharing within the class repository.
set.seed(123) 
hrs_sample <- hrs_long %>%
  slice_sample(n = 1000)

# 3. save to processed-data 
saveRDS(hrs_sample, here("data", "processed-data", "randhrs_sample_1000.rds"))
```


# Check data 
First we can look at the data. For the main HRS data, we have 280,343 observations about 42405 unique individuals (HHIDPN) across 16 waves. The large number of observations is due to the longitudinal structure of HRS, where each person can have multiple rows corresponding to different survey waves. 

```{r}
glimpse(hrs_long)
head(hrs_long)
hrs_long %>% 
  summarise(total_people = n_distinct(HHIDPN))
```

```{r}
glimpse(lange_weir_raw)
head(lange_weir_raw)
```

# Cleaning

Next, follow the following steps to process the data:

1. Reshape the data of lange_weir from wide to long format.
```{r}
# Variable stubs (without year suffix)
cog_stubs <- c("bwc20_imp", "cogfunction", "cogivewmode", "cogtot27_imp",
               "dlrc_imp", "fbwc20_imp", "fdlrc_imp", "fimparate_imp",
               "fimrc_imp", "fmemoryp_imp", "fnumiadl_imp", "fser7_imp",
               "imparate_imp", "imrc_imp", "interview",
               "memoryp_imp", "numiadl_imp", "proxy", "prxyscore_imp", "ser7_imp")    

# reshape lange weir data from wide to long format
lange_weir_long <- lange_weir_raw %>%
  pivot_longer(
    cols = matches(paste0(cog_stubs, collapse = "|")),
    names_to = c(".value", "studyyr"),
    names_pattern = paste0("^(", paste(cog_stubs, collapse = "|"), ")(\\d{4})$")
  ) %>%
  mutate(studyyr = as.integer(studyyr)) %>%
  # Keep only those interviewed in that wave (same as SAS/Stata code)
  filter(interview == 1) %>%
  arrange(hhid, pn, studyyr)
```

2. Merge the lange weir data with the main HRS data. Before merging, we need to make sure that the key variables for merging (HHIDPN and studyyr) are in the same format in both datasets. In the main HRS data, we have HHIDPN and STUDYYR variables, while in the lange weir data, we have hhid, pn, and studyyr. And we should limit the year from 1996 to 2020 to match the waves in the main HRS data to make sure consistency.  
We will merge them by person (HHIDPN) and year.
```{r}
# limit the year from 1996 to 2020 to match the waves in the two datasets
# Remove .y columns (keep .x as primary)
hrs_long_filtered <- hrs_long %>%
  filter(STUDYYR >= 1996 & STUDYYR <= 2020)
lange_weir_filtered <- lange_weir_long %>%
  filter(studyyr >= 1996 & studyyr <= 2020)

check_n(hrs_long_filtered)
```
```{r}
# creat HHIDPN variable in lange weir data 
lange_weir_filtered <- lange_weir_filtered %>%
  mutate(HHIDPN = as.numeric(paste0(hhid, pn)),
         STUDYYR = studyyr) %>%   # Rename to match HRS column name
  select(-hhid, -pn, -studyyr)    # Drop old ID columns
```
```{r}
# merge the two datasets
hrs_cog <- hrs_long_filtered %>%
  left_join(lange_weir_filtered, by = c("HHIDPN", "STUDYYR"))
check_n(hrs_cog)
```

3. Select the variables needed for the analysis.
```{r}
# ── Select only variables needed based on codebook ────────────────────────
hrs_cog_vars <- hrs_cog %>%
  select(
    # Key ID and wave variables
    HHIDPN, STUDYYR, WAVE_NUMBER,
    # Time-invariant demographics (from rwide)
    RABYEAR,       # Birth year
    RARACEM,       # Race
    RAHISPAN,      # Ethnicity
    RAEDUC,        # Education
    RAGENDER,      # Gender
    RAWTSAMP,      # Sampling weight
   
   # Respondent wave-varying health variables (from rlong)
    R_MSTAT,       # Marital status
    R_BMI,         # BMI (self-reported)
    R_PMBMI,       # BMI (physically measured, 2006-2018)
    R_HIBP,        # Hypertension
    R_DIAB,        # Diabetes
    R_CANCR,       # Cancer
    R_LUNG,        # Chronic lung disease
    R_HEART,       # Heart disease
    R_STROK,       # Stroke
    R_DEPRES,      # Depression (CESD)
    R_SMOKEV,      # Ever smoked
    R_SMOKEN,      # Smokes now
    R_DRINKD,      # Days/week drinks (1996-2020)
    R_DRINKN,      # Drinks/day
    R_LBRF,        # Labor force status
    R_HIGOV,       # Covered by government health plan
    R_GOVMR,       # Medicare
    R_GOVMD,       # Medicaid
    R_GOVVA,       # Champus/VA
    R_GOVOT,       # Other government plan
    R_PRPCNT,      # Number of private health insurance plans
    R_COVR,        # Covered by employer health insurance
    R_COVS,        # Covered by spouse employer health insurance
    R_HIOTHP,      # Other health insurance
    R_HILTC,       # Long-term care insurance
    
    # Household wave-varying variables (from hlong)
    H_HHRES,       # Number of people in household
    H_ATOTB,       # Net total wealth
    H_ITOT,        # Total household income

    # Cognitive variables (from lange-weir)
    cogfunction,   # Outcome: cognition category (1=Normal, 2=CIND, 3=Demented)
    cogtot27_imp,  # 27-point cognitive total score
    cogivewmode,   # Interview mode
    proxy,         # Proxy respondent type
    imrc_imp,      # Immediate word recall
    dlrc_imp,      # Delayed word recall
    ser7_imp,      # Serial 7s
    bwc20_imp,     # Backwards count
  )

check_n(hrs_cog_vars)
```

4. Check missing values for each variable
```{r}
# Check missing values for each variable
missing_summary <- hrs_cog_vars %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_missing") %>%
  mutate(pct_missing = round(n_missing / nrow(hrs_cog_vars) * 100, 1)) %>%
  arrange(desc(pct_missing))
print(missing_summary)
# Delete variables with more than 50% missing values 
hrs_cog_miss <- hrs_cog_vars %>%
  select(-c(R_GOVOT, R_PMBMI, RAWTSAMP, cogivewmode)) 
```

5. Create demographic, socioeconomic, health, and BMI-related covariates. 

```{r}
hrs_cog_cov <- hrs_cog_miss %>%
  mutate(
    # Age at interview
    AGE = STUDYYR - RABYEAR,

    # Create HAS_ANY_INSURANCE variable (0/1/NA)
    # 1 = has any insurance (Medicare, Medicaid, VA, or any private plan), 0 = no insurance
    # NA = all insurance variables are missing
    HAS_INS = case_when(
      R_GOVMR == 1 | R_GOVMD == 1 | R_GOVVA == 1 |
        (R_PRPCNT >= 1 & !is.na(R_PRPCNT)) ~ 1,
      is.na(R_GOVMR) & is.na(R_GOVMD) & is.na(R_GOVVA) & is.na(R_PRPCNT) ~ NA_real_,
      TRUE ~ 0
    ) %>%
      factor(levels = 0:1, labels = c("No insurance", "Has insurance")),

    # Race/ethnicity 3 categories: Non-Hispanic White, Non-Hispanic Black, Hispanic or Other
    RACE_ETH3 = case_when(
      RAHISPAN == 0 & RARACEM == 1 ~ 1,
      RAHISPAN == 0 & RARACEM == 2 ~ 2,
      RAHISPAN == 0 & RARACEM == 3 ~ 3,
      RAHISPAN == 1 ~ 3,
      TRUE ~ NA_real_
    ) %>%
      factor(
        levels = 1:3,
        labels = c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic or Other")
      ),

    # Education categories: Less than high school, GED, High school graduate, College or above (merge some categories based on distribution and codebook)
    RAEDUC = case_when(
      RAEDUC %in% c(4, 5) ~ 4,
      TRUE ~ RAEDUC
    ) %>%
      factor(
        levels = 1:4,
        labels = c("Less than high school", "GED", "High school graduate", "College or above")
      ),
      
    # Gender
    RAGENDER = factor(RAGENDER, levels = 1:2, labels = c("Male", "Female")),

    # Smoking status: Never smoker, Ever/Current smoker, NA=Missing/Unknown
    SMOKESTATUS = case_when(
      R_SMOKEV == 0 ~ 0,
      R_SMOKEV == 1 ~ 1,
      TRUE ~ NA_real_
    ) %>%
      factor(levels = 0:1, labels = c("Never smoker", "Ever/Current smoker")),

    # Employment status: Employed (FT/PT), Unemployed/Not in labor force, Retired, Disabled
    EMP_STATUS = case_when(
      R_LBRF %in% c(1, 2) ~ 1,
      R_LBRF %in% c(3, 7) ~ 2,
      R_LBRF %in% c(4, 5) ~ 3,
      R_LBRF == 6 ~ 4,
      TRUE ~ NA_real_
    ) %>%
      factor(
        levels = 1:4,
        labels = c("Employed (FT/PT)", "Unemployed/Not in labor force", "Retired", "Disabled")
      ),

    # Income and wealth variables: Log transform total income and wealth to reduce skewness
    LN_HITOT = log(H_ITOT + 1),
    LN_WEALTH = asinh(H_ATOTB),

    # Depression: Recode R_DEPRES to factor with labels
    R_DEPRES = factor(R_DEPRES, levels = 0:1, labels = c("No", "Yes")),

    # Cognition outcome: convert to factor
    cogfunction = factor(cogfunction,
                         levels = 1:3,
                         labels = c("Normal", "CIND", "Demented")),
    # BMI category based on R_BMI: Thin (<18.5), Normal (18.5-24.9), Overweight (25.0-29.9), Obese (>=30.0)
    BMI_CAT = case_when(
      R_BMI < 18.5 ~ 1,
      R_BMI >= 18.5 & R_BMI < 25 ~ 2,
      R_BMI >= 25 & R_BMI < 30 ~ 3,
      R_BMI >= 30 ~ 4,
      TRUE ~ NA_real_
    ) %>%
      factor(
        levels = 1:4,
        labels = c("Thin (<18.5)", "Normal (18.5-24.9)", "Overweight (25.0-29.9)", "Obese (>=30.0)")
      )
  ) %>%
  mutate(
  # Recode chronic condition variables to binary (1=Yes, 0=No/other), then count number of chronic conditions.
    across(
      c(R_HIBP, R_DIAB, R_CANCR, R_LUNG, R_HEART, R_STROK),
      ~ case_when(
        . %in% c(1, 3, 6) ~ 1,
        . %in% c(0, 4) ~ 0,
        . == 5 ~ NA_real_,
        TRUE ~ NA_real_
      ),
      .names = "{.col}_bin"
    )
  ) %>%
  mutate(
    CHRONIC_N = rowSums(select(., ends_with("_bin")), na.rm = TRUE)
  )

```

6. Create baseline covariates from the 1996 wave and merge them back to the person-wave data.

```{r}
baseline_1996 <- hrs_cog_cov %>%
  filter(STUDYYR == 1996) %>%
  transmute(
    HHIDPN,
    BASELINE_AGE = AGE,
    BASELINE_BMI = R_BMI,
    BASELINE_BMI_CAT = BMI_CAT
  )
hrs_cog_base <- hrs_cog_cov %>%
  left_join(baseline_1996, by = "HHIDPN")
```
7. Restrict to the Baseline-Eligible Cohort
```{r}
# Restrict to respondents with valid baseline information and age 50+ at baseline
hrs_base_cohort <- hrs_cog_base %>%
  filter(
    !is.na(BASELINE_AGE),
    !is.na(BASELINE_BMI),
    BASELINE_AGE >= 50
  )

check_n(hrs_base_cohort)
```

# Select Final Variables for the Clean Longitudinal Master File
At this point, we select only the variables needed for the regression analysis (may edit later).

```{r}
hrs_long_clean <- hrs_base_cohort %>%
  rename_with(toupper) %>%
  select(
    HHIDPN, STUDYYR, WAVE_NUMBER,

    # Cognition outcomes
    COGFUNCTION,
    COGTOT27_IMP,

    # BMI variables
    R_BMI,
    BMI_CAT,
    BASELINE_AGE,
    BASELINE_BMI,
    BASELINE_BMI_CAT,
    AGE,

    # Demographic and socioeconomic covariates
    RAGENDER,
    RACE_ETH3,
    RAEDUC,
    LN_WEALTH,
    LN_HITOT,

    # Health-related covariates
    CHRONIC_N,
    HAS_INS,
    EMP_STATUS,
    SMOKESTATUS,
    R_DEPRES
  )

check_n(hrs_long_clean)
```


# Create exposure variable
We create BMI varibility and mean BMI across waves for each person, and categorize BMI variability into tertiles based on the wave 9(the longest follow-up wave). 

```{r}
# impute BMI for each person using linear interpolation across waves.  
hrs_long_clean <- hrs_long_clean %>%
  group_by(HHIDPN) %>%
  arrange(HHIDPN, STUDYYR) %>%
  mutate(
    R_BMI_IMP = na.approx(R_BMI, 
                                x = STUDYYR,      # Use study year as x-axis
                                na.rm = FALSE,    # Keep NA if cannot interpolate (e.g. at edges)
                                rule = 2)         # rule=2: extrapolate at edges using nearest value
  ) %>%
  ungroup()

hrs_long_clean <- hrs_long_clean %>%
  group_by(HHIDPN) %>%
  filter(!any(is.na(R_BMI_IMP))) %>%
  ungroup()
check_n(hrs_long_clean)
```


```{r}
# Define wave windows (starting from 1996)
wave_years <- list(
  w3 = c(1996, 1998, 2000),
  w4 = c(1996, 1998, 2000, 2002),
  w5 = c(1996, 1998, 2000, 2002, 2004),
  w6 = c(1996, 1998, 2000, 2002, 2004, 2006),
  w7 = c(1996, 1998, 2000, 2002, 2004, 2006, 2008),
  w8 = c(1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010),
  w9 = c(1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012)
)

# Calculate BMI variability for each wave window using imputed BMI
bmi_variability_list <- list()

for (w in names(wave_years)) {
  years <- wave_years[[w]]
  
  bmi_variability_list[[w]] <- hrs_long_clean %>%
    filter(STUDYYR %in% years) %>%
    group_by(HHIDPN) %>%
    arrange(HHIDPN, STUDYYR) %>%
    summarise(
      n_obs    = sum(!is.na(R_BMI_IMP)),
      BMI_MEAN = mean(R_BMI_IMP, na.rm = TRUE),
      BMI_SD   = sd(R_BMI_IMP, na.rm = TRUE),
      BMI_CV   = BMI_SD / BMI_MEAN * 100,
      BMI_ARV  = mean(abs(diff(R_BMI_IMP[!is.na(R_BMI_IMP)])), na.rm = TRUE),
      .groups  = "drop"
    ) %>%
    rename_with(~ paste0(., "_", toupper(w)), -HHIDPN)
}

# Merge all wave windows into one wide table
bmi_variability <- bmi_variability_list %>%
  reduce(full_join, by = "HHIDPN")

# Create BMIV_CAT for each wave window based on W9 tertile cutpoints
w9_tertile_breaks <- quantile(bmi_variability$BMI_CV_W9,
                              probs = c(0, 1/3, 2/3, 1),
                              na.rm = TRUE)

cat("W9 tertile cutpoints:\n")
print(w9_tertile_breaks)

for (w in names(wave_years)) {
  cv_col  <- paste0("BMI_CV_", toupper(w))
  cat_col <- paste0("BMIV_CAT_", toupper(w))
  
  bmi_variability <- bmi_variability %>%
    mutate(!!cat_col := cut(.data[[cv_col]],
                            breaks = w9_tertile_breaks,
                            labels = c("Low variability",
                                       "Medium variability",
                                       "High variability"),
                            include.lowest = TRUE))
}

# Merge back to long table
hrs_final <- hrs_long_clean %>%
  left_join(bmi_variability, by = "HHIDPN")

cat("Final dimensions:", dim(hrs_final), "\n")
check_n(hrs_final)

```

# Save the cleaned and processed data for analysis
```{r}
saveRDS(hrs_final, here("data", "processed-data", "hrs_final_long.rds"))
```













